The algebraic hierarchy in type theory using type classes

### 2010-06-22 jar.tex
Add subsection about implicit syntax-directed algorithm composition (to establish the Decision class).

\section{Predicate classes}

@@

@@

However, there are contexts (in other proofs) where \lstinline|Prop|-sorted construction of subalgebras really \emph{is} appropriate. Unfortunately, Coq's universe polymorphism is not yet up to the task of letting us use a single set of definitions to handle both cases. In particular, there is no universe polymorphism for definitions (as opposed to inductive definitions) yet. We will return to this point later. In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in about a hundred lines of duplication.

@@

  Theorem first_iso: iso_arrows back forth.
\end{lstlisting}
The \lstinline|Program| command generates proof obligations (not shown) expressing that these two arrows are indeed homomorphisms. The theorem itself is trivial.

@@

For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour. As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations. Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic representation in the equational theory of semirings. We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in~\ref{quoting}.

Thanks to our close integration of universal algebra we can actually obtain a \lstinline|Naturals| implementation completely automatically, by invoking a generic construction of initial models built from the closed term algebra for the signature along with a setoid equality expressing the congruence closure of the identities in the equational theory. However, this implementation is very useful neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted), because for example deciding equality for the aforementioned setoid equality is far harder than for, say, \lstinline|nat|, because it requires a normalization procedure.

@@

The generic \lstinline|Decision| instance for \lstinline|Naturals| equality implemented by mapping to \lstinline|nat| will typically be far less efficient than a specialized implementation for a particular representation of the natural numbers. Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist, and we can even deprioritize the generic instance so that instance resolution will always pick the specialization when the representation is known.

To permit a \emph{generic} function operating on naturals to take advantage of specialized operations, we simply introduce an additional instance parameter:
\begin{lstlisting}
Definition calculate_things `{Naturals N} `{!forall n m: N, Decision (n = m)}
  (a b: nat): ... := ... decide (a = b) ... .
\end{lstlisting}
Without the \lstinline|Decision| parameter \lstinline|calculate_things| would be equally correct, but could be less efficient. Thus, by this scheme one can start by writing correct-but-possibly-inefficient programs that make use of generic operation instances, and then selectively improve efficiency of key algorithms simply by adding additional operational type class instance parameters where profiling shows it to make a significant difference, without changing their definition body.

>>>>

\section{Predicate classes and operational classes}

@@

\subsection{Implicit syntax-directed algorithm composition}

Before we proceed with a discussion of the algebraic hierarchy based on type classes, we briefly highlight one specific operational type class because we will use it later and because it is a particularly nice illustration of another neat application of operational type classes. The operation in question is that of deciding a proposition:
\begin{lstlisting}
  Class Decision (P: Prop): Type := decide: sumbool P (¬ P).
\end{lstlisting}
Here, \lstinline|sumbool| is just the (informative) sum of proofs.

\lstinline|Decision| is a very general purpose type class that also works for predicates. For instance, to declare a parameter expressing decidability of, say, (setoid) equality on a type \lstinline|X|, we write: \lstinline|`{Π a b: X, Decision (a = b)}|. To then use this (unnamed) decider to decide a particular equality, we simply say \lstinline|decide (x = y)|, and instance resolution will resolve the decider we declared.

With \lstinline|Decision| as a type class, we can very easily define composite deciders for things like conjunctions and quantifications over (finite) domains:
\begin{lstlisting}
  Instance decide_conj `{Decision P} `{Decision Q}: Decision (P ∧ Q).
\end{lstlisting}
With these in place, we can just say \lstinline|decide (x = y ∧ p = q)| and let instance resolution automatically compose a decision procedure that can decide the specified proposition. This style of syntax-directed implicit composition of algorithms is very convenient and highly expressive.

@@

However, there are contexts (in other proofs) where \lstinline|Prop|-sorted construction of subalgebras really \emph{is} appropriate. Unfortunately, Coq's universe polymorphism is not yet up to the task of letting us use a single set of definitions to handle both cases. In particular, there is no universe polymorphism for definitions (as opposed to inductive definitions) yet. We will return to this point later. In our development, we have two sets of definitions, one for \lstinline|Prop| and one for \lstinline|Type|, resulting in duplication of about a hundred lines of code.

@@

  Theorem first_iso: iso_arrows back forth.
\end{lstlisting}
The \lstinline|Program| command generates proof obligations (not shown) expressing that these two arrows are indeed homomorphisms. The proof of the theorem itself is trivial.

@@

For more complex properties such as injectivity of partially applied addition and multiplication, we take a longer detour. As part of our universal algebra theory, we have proved that proofs of statements in the language of an equational theory can be transferred between isomorphic implementations. Combined with the trivially proved property that any two initial objects in a category are isomorphic, this lets us tranfer proofs of such statements between implementations of \lstinline|Naturals|, requiring only that we reflect the concrete statement (expressed in terms of the operational type classes) to a symbolic statement in the language of semirings. We intend to eventually make this reflection completely automatic using type class based quotation techniques along the lines of those described in~\ref{quoting}.

Thanks to our close integration of universal algebra we can actually obtain a \lstinline|Naturals| implementation completely automatically, by invoking a generic construction of initial models built from the closed term algebra for the signature along with a setoid equality expressing the congruence closure of the identities in the equational theory. However, this implementation is very useful neither in terms of efficiency, nor as a canonical implementation (to be used as the basis for theory and programs that are then subsequently lifted), because for example deciding equality for the aforementioned setoid equality is far harder than for, say, \lstinline|nat| (because it requires a normalization procedure).

@@

% todo: decision
The generic \lstinline|Decision| instance for \lstinline|Naturals| equality implemented by mapping to \lstinline|nat| will typically be far less efficient than a specialized implementation for a particular representation of the natural numbers. Fortunately, with Coq's type classes it is no problem for instances overlapping in this way to co-exist, and we can even deprioritize the generic instance so that instance resolution will always pick the specialization when the representation is known.

To permit a \emph{generic} function operating on naturals to take advantage of specialized operations, we simply introduce an additional instance parameter:
\begin{lstlisting}
Definition calculate_things `{Naturals N} `{!Π n m: N, Decision (n = m)}
  (a b: nat): ... := ... decide (a = b) ... .
\end{lstlisting}
Without the \lstinline|Decision| parameter \lstinline|calculate_things| would be equally correct, but could be less efficient. Thus, by this scheme one can start by writing correct-but-possibly-inefficient programs that make use of generic operation instances, and then selectively improve efficiency of key algorithms simply by adding additional operational type class instance parameters where profiling shows it to make a significant difference, without changing their definition body.
